Scaffolder - Software for Microbial Genome Scaffolding.

### 2011-02-28 src/article.tex
Christina changes to introduction

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each sequence represents the machine-read
nucleotides from a short region in the genome. At the time of writing, high
throughput sequencing is limited to producing reads less than 1000 nucleotides
in length. Therefore to resolve a sequence longer than this, such as a complete
genome, these numerous smaller fragments must be pieced together. The process
of joining reads into longer sequences is the `assembly' stage of a genome
project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle.
Each nucleotide read represents a single piece and the final genome sequence is
the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to not
enough or multiple overlaps between sequence reads and is analogous to missing
pieces in the jigsaw or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has lead
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}; genome assembly software may however be unable
to produce a complete sequence from these reads. In many cases assembly
algorithms will generate several large contiguous regions of sequence
(`contigs') composed of the many available individual reads, which represent the
assembly of reads into the longest possible sequences given the data.
Nonetheless, these contigs represent a fragmented picture of the genome and
require additional work to join together into larger sequence lengths. \pb

@@

Scaffolding is the process of joining a series disconnected contigs into
complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in joining a greater subset of contigs together or resolving
gaps. An overview of required steps in the scaffolding process is outlined
below.

@@

During the scaffolding process closing and filling gaps between contigs
improves the completeness of the genome scaffold. Closing gaps may require
returning to the laboratory to perform additional sequencing or using
computational methods of estimating the unknown sequence. This additional
sequence is used to replace the gap between two contigs joining them into
a single sequence if they are not already scaffolded together. Once all contigs
have been joined and gaps in a scaffold closed, the genome may be considered
finished.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use paired-read data, where the
sequencing reads are produced in pairs at a known distance apart in the genome.
The occurrence of a pair of reads in separate contigs can be used to estimate
the order and distance between the contigs. Alternatively, laboratory methods
may use PCR to amplify the unknown DNA in a gap region, then use traditional
Sanger sequencing to determine the gap sequence. Computational methods are more
preferable as they are less costly in laboratory time and materials compared to
manual gap resolution and uses the available data that exists following
sequencing \cite{nagarajan2010}. \pb

Synteny methods compare the assembled contigs to a more complete reference
genome sequence searching for areas of sequence similarity between the two.
Areas of corresponding sequence between the two sets are can be used to infer
contig placement and build the scaffold \cite{richter2007,zhao2008,assefa2009}.
Recombination between the two genomes can however reduce the effectiveness of
this analysis. Repeat regions may be responsible for many of the gaps in
building a genome sequence. The existence of tandemly repeated nucleotide
regions produces many similar reads in the sequence process. The assembly of
reads generated from repeat regions can result in the assemble collapsing into
a single repeat instance or ignored by conservative assembly algorithms. These
regions may be resolved by using reassembly algorithms specifically for repeat
resolution \cite{mulyukov2002,koren2010}. Similarly a related approach may use
unassembled sequence reads matching the regions around a scaffold gap to
construct a uniquely overlapping set of reads across the region
\cite{tsai2010}. \pb

Paired-read data provides an extra level of information on how contigs may be
scaffolded together. Heuristic algorithms may take advantage of this data to
search for optimal configuration of contigs in the scaffold using paired-read
distances \cite{dayarian2010,boetzer2011}. Additional data using a reference
genome may also be combined when estimating the best contig configuration
\cite{pop2004}. Finally when the scaffold cannot be completely resolved
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions
\cite{gordon2001,nagarajan2010}. \pb

@@

The software outlined here ``Scaffolder'' is aimed at creating a syntax and
software framework for producing and editing a genome scaffold. Scaffolder uses
a specific file format to define how contigs are joined, additional sequences
inserted, and for the specification of unknown regions. This syntax allows
a scaffold to be updated by simply editing the scaffold file and allows the
scaffolding process to be both reproducible and deterministic. \pb

>>>>

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each sequence represents the machine-read
nucleotides from a short region in the genome. At the time of writing, high
throughput sequencing is limited to producing reads less than 1,000 nucleotides
in length. Therefore to resolve a sequence longer than this, such as a complete
genome, these numerous smaller fragments must be pieced together. The process
of joining reads into longer sequences is the `assembly' stage of a genome
project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle.
Each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to not
enough or multiple overlaps between sequence reads and is analogous to missing
pieces in the jigsaw or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has led
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}; genome assembly software may however be unable
to produce a complete sequence from these reads. In many cases assembly
algorithms will generate several large contiguous regions of sequence
(`contigs') composed of the many available individual reads, which represent
the assembly of reads into the longest possible sequences given the data.
Nonetheless, these contigs represent a fragmented picture of the genome and
require additional work to join together into larger sequence lengths. \pb

@@

Scaffolding is the process of joining a series of disconnected contigs into
a complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in joining a greater subset of contigs together or resolving
gaps. An overview of required steps in the scaffolding process is outlined
below.

@@

During the scaffolding process, closing and filling gaps between contigs
improves the completeness of the genome scaffold. Closing gaps may require
returning to the laboratory to perform additional sequencing or using
computational methods of estimating the unknown sequence. This additional
sequence is used to replace the gap between two contigs joining them into
a single sequence if they are not already scaffolded together. Once all contigs
have been joined and gaps in a scaffold closed, the genome may be considered
finished.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use paired-read data, where the
sequencing reads are produced in pairs at a known distance apart in the genome.
The occurrence of a pair of reads in separate contigs can be used to estimate
the order and distance between the contigs. Alternatively, laboratory methods
may use PCR to amplify the unknown DNA in a gap region, then use traditional
Sanger sequencing to determine the gap sequence. Computational methods are more
preferable, as they are less costly in laboratory time and materials compared
to manual gap resolution and uses the available data that exists following
sequencing \cite{nagarajan2010}. \pb

Synteny methods compare the assembled contigs to a more complete reference
genome sequence searching for areas of sequence similarity between the two.
Areas of corresponding sequence between the two sets can be used to infer
contig placement and to build the scaffold
\cite{richter2007,zhao2008,assefa2009}. Recombination between the two genomes
can, however, reduce the effectiveness of this analysis. Repeat regions may be
responsible for many of the gaps in building a genome sequence. The existence
of tandemly repeated nucleotide regions produces many similar reads in the
sequence process. The assembly of reads generated from repeat regions can
result in the assemble collapsing into a single repeat instance or ignored by
conservative assembly algorithms. These regions may be resolved by using
reassembly algorithms specifically for repeat resolution
\cite{mulyukov2002,koren2010}. Similarly a related approach may use unassembled
sequence reads matching the regions around a scaffold gap to construct
a uniquely overlapping set of reads across the region \cite{tsai2010}. \pb

Paired-read data provides an extra level of information about how contigs may
be scaffolded together. Heuristic algorithms may take advantage of this data to
search for optimal configuration of contigs in the scaffold using paired-read
distances \cite{dayarian2010,boetzer2011}. Additional data using a reference
genome may also be combined when estimating the best contig configuration
\cite{pop2004}. Finally when the scaffold cannot be completely resolved,
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions
\cite{gordon2001,nagarajan2010}. \pb

@@

The software outlined here, ``Scaffolder'', is aimed at creating a syntax and
software framework for producing and editing a genome scaffold. Scaffolder uses
a specific file format to define how contigs are joined, additional sequences
inserted, and for the specification of unknown regions. This syntax allows
a scaffold to be updated by simply editing the scaffold file and allows the
scaffolding process to be both reproducible and deterministic. \pb
